Global land characterisation using cover fractions at 100m resolution

### 2020-08-05 article/article.tex
Draft 3

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lineno}
\usepackage{microtype}
\usepackage{amsmath}
\usepackage{indentfirst}
\usepackage{booktabs} % For \toprule, etc.
%\usepackage[sort&compress,sectionbib]{natbib} % already in elsarticle
\usepackage{graphicx}
\usepackage{adjustbox} % For resizing tables
\usepackage{array} % For mid-alignment (m{xcm}) in tables
\usepackage{rotating} % Sideways figures
%\bibliographystyle{apalike} % elsarticle defined at the top

@@

% My own definition for the sections that start with a bold name but don't appear in TOC
\newcommand{\minisection}[1]{\paragraph{#1}}%{\medskip \textbf{#1:}}

@@

% Acronyms - cite with \ac{}
\newacronym{GNSS}{GNSS}{Global Navigation Satellite System}
\newacronym{OA}{OA}{overall accuracy}
\newacronym{UA}{UA}{user accuracy}
\newacronym{PA}{PA}{producer accuracy}
\newacronym{RMSE}{RMSE}{root mean squared error}
\newacronym{MAE}{MAE}{mean absolute error}
\newacronym{ME}{ME}{mean error}
\newacronym{NSE}{NSE}{Nash–Sutcliffe model efficiency coefficient}
\newacronym{SDG}{SDG}{Sustainable Development Goal}
\newacronym{LC-CCI}{LC-CCI}{ESA Climate Change Initiative Land Cover}
\newacronym{CGLS-LC100}{CGLS-LC100}{Copernicus Global Land Services 100 m Land Cover}
%\newacronym{CGLS}{CGLS}{Copernicus Global Land Services}
\newacronym{FROM-GLC}{FROM-GLC10}{Finer Resolution Observation and Monitoring Global Land Cover}
\newacronym{IIASA}{IIASA}{International Institute for Applied Systems Analysis}
\newacronym{WUR}{WUR}{Wageningen University \& Research}
\newacronym{SMA}{SMA}{spectral mixture analysis}
\newacronym{SVM}{SVM}{support vector machine}
\newacronym{MLP}{MLP}{multi-layer perceptron}
\newacronym{NN}{NN}{neural network}
\newacronym{LOESS}{LOESS}{locally estimated scatterplot smoothing}
\newacronym[firstplural={vegetation indices (VIs)}]{VI}{VI}{vegetation index}
\newacronym{GLSDEM}{GLSDEM}{Global Land Survey Digital Elevation Model}
\newacronym{TPI}{TPI}{Terrain Position Index}
\newacronym{OSAVI}{OSAVI}{Optimised Soil-Adjusted Vegetation Index}
\newacronym{NDVI}{NDVI}{Normalised Difference Vegetation Index}
\newacronym{NDMI}{NDMI}{Normalised Difference Moisture Index}
\newacronym{NIRv}{NIR\textsubscript{v}}{\gls{NIR} of vegetation}
\newacronym{EVI}{EVI}{Enhanced Vegetation Index}
\newacronym{NIR}{NIR}{near infra-red}
\newacronym{SCM}{SCM}{subpixel confusion-uncertainty matrix}
\newacronym{FNC}{FNC}{fuzzy nearest centroid}
\newacronym{PLS}{PLS}{partial least squares}
\newacronym{MLR}{MLR}{multinomial logistic regression}
\newacronym{RF}{RF}{random forest}
\newacronym{TOC}{TOC}{Top-of-Canopy}
\newacronym{CART}{CART}{classification and regression tree}
\newacronym{GLM}{GLM}{General linear regression model}
\newacronym{LCCS}{LCCS}{UN Land Cover Classification Scheme}
\newacronym{GDAL}{GDAL}{Geospatial Data Abstraction Library}
%\makenoidxglossaries
% Disable abbreviating RF for now [text={random forest}]
%\glsunset{RF}

@@

\begin{abstract} % elsarticle: environment instead of a command \abstract{
Currently most global land cover maps are produced with discrete classes, which express the dominant land cover class in each pixel, or a combination of several classes at a particular ratio. In contrast, land cover fraction mapping allows for expressing the proportion of each pure class in each pixel, which increases precision and reduces legend complexity. To map land cover fractions, regression rather than classification algorithms are needed, and multiple approaches are available for this task.

@@

One challenge for land cover fraction mapping models is data sparsity: the dataset is zero-inflated, but regression favours the mean, thus 0\% and 100\% are difficult to predict well. We proposed a new solution by combining three models. A binary model determines whether a pixel is pure; if so, it is processed using a classification model; otherwise with a regression model.

@@

\minisection{Highlights}

\begin{itemize}
    \item 3 machine learning model groups tested for deriving 7 land cover fractions globally
    \item Random forest regression performed the best (17.3 RMSE or 7.9 MAE)
    \item Optical remote sensing data was the most important input for model predictions
    \item Independent validation showed highest accuracies for built-up and water fractions
    \item Multi-fraction land cover mapping shown as an alternative to discrete classification
\end{itemize}

%\pagebreak
%\tableofcontents
%\pagebreak

@@

Land cover, as one of the key variables for monitoring a number of \glspl{SDG}, has received increasing amounts of attention, also as new satellites are launched and the capacity for land monitoring increases.
In this context, new global land cover maps have emerged to better map the current land cover of the world, as well as to track land cover change.
Some of the recent achievements have been the \ac{LC-CCI} map \citep{defourny2012cci} that aims at producing a long-term set of consistent global annual medium-resolution land cover maps aimed at the climate community, \glsentryfull{CGLS-LC100} map \citep{buchhorn_moderate_2019, buchhorn_copernicus_2020} that aims at a finer spatial resolution and high quality with yearly updates since 2015, and the \ac{FROM-GLC} map \citep{fromglc2019} that showcases the potential of land cover mapping at 10 m resolution.

Except for the \ac{CGLS-LC100} product with its cover fraction layers, all other global land cover maps, including previous maps such as by \citet{bartholome2005glc2000, friedl2010modis, arino2007globcover, see2015hybrid, chen2015globeland30}, are provided with discrete classes (also known as ``hard'' or ``crisp'' classification), where each pixel of the map can only represent a single class.
However, this results in less precise estimation of land cover due to mixed pixels, where in reality there are multiple land cover classes in the area covered by a single pixel.
This issue becomes worse at coarse resolutions and heterogeneous landscapes.
It may result in biases, for instance, a sparse forest may be classified as grassland, ignoring the relatively few trees in the area, and thus ultimately lead to an underestimation of tree cover at large scales.

A potential solution to this issue is land cover fraction mapping, where instead of a discrete class, the proportion of every class in the legend is reported for every pixel of the map.
This is also called ``fuzzy'' or ``soft'' classification, and sometimes ``subpixel'' classification or ``linear mixture modelling'' \citep{Okeke2006fuzzyexponent}.
Related terms are ``super-resolution mapping'' and ``subpixel mapping'', but this concept relates also to the allocation of land cover fractions inside a pixel spatially, rather than just determining the fraction ratio per pixel.

Land cover fraction mapping has been attempted in the past, both in research \citep{adams_classification_1995, foody1996fuzzyevaluation, colditz_land_2011, sharma_assessing_2011, uma_shankar_wavelet-fuzzy_2011, dwivedi_optimisation_2012, lizarazo_quantitative_2012, gessner_estimating_2013, okujeni_generalizing_2018} and production (e.g. \citealp{hansen2000hardtree, hansen_continuous_2011, pengra_global_2015, hansen_global_2003, sexton_global_2013}).
Nevertheless, to date only \ac{CGLS-LC100} \citep{marcel_buchhorn_copernicus_2019} provides global maps with fractions of every major land cover class \citep{nandin-erdene_tsendbazar_copernicus_2019}.

Land cover fraction mapping can be performed using a variety of different approaches and algorithms.
In its core, it is a regression rather than a classification problem, as the output is a fraction of a label rather than a label itself.
Popular approaches include fuzzy nearest centroid regression \citep{zhang2001fullyfuzzy}, \ac{SMA} (e.g. \citealp{shimabukuro1991least, adams_classification_1995, hobbs2003linear, yang_landsat_2012}), \ac{RF} regression \citep{walton2008subpixelrf}, \gls{SVM} regression \citep{walton2008subpixelrf}, Cubist regression \citep{walton2008subpixelrf}, \gls{MLP} \glspl{NN} \citep{zhang2001fullyfuzzy}, genetic algorithms \citep{stavrakoudis_boosted_2011} and wavelet transformation \citep{uma_shankar_wavelet-fuzzy_2011}.
So far other studies have only compared a few of these methods at once, and never with an exhaustive set of land cover classes nor at global scale.

In addition to classification methods, one aspect that is important for classification accuracy is the data (covariates) that the algorithm is trained on and predict land cover from \citep{yu2014metadiscoveries}. In addition to optical covariates (blue, red, near infrared, shortwave infrared spectral bands), temporal covariates (growing season start, duration, growing intensity) are commonly used to improve the separability of crops \citep{jakubauskas2001harmonic}. Elevation covariates (elevation, slope, aspect) are known to improve the separability between tree classes \citep{burrough2001fuzzy}. \Glspl{VI} condense data and reduce the noise inherent in single bands, and thus are often used to improve land cover class separability, e.g. for wetlands \citep{sader1995wetlands}, as well as for defining classification rules in rule-based fuzzy classification \citep{baraldi2006rulebased}. Therefore in this paper such covariates have also been used in order to increase the accuracy of land cover fraction regression model predictions, and their importance for the predictions was analysed.

Lastly, accuracy assessment by itself is often times challenging, especially at a global scale.
It requires a comprehensive dataset across the globe that would be comparable with the modelled land cover, and yet independent of the training data used to create it.
For land cover fraction mapping it is even more challenging, as fraction information needs to be available, or alternatively fine spatial resolution data from which the fractions can be calculated.
Because of these challenges, the information about the accuracy of the existing global land cover products is often limited, which in turn is problematic for the users to decide on what product suits their needs the best.
Users would also benefit from knowing the spatial variation of accuracy, as models may be more accurate at certain locations of the world and less so at others.
In this study, we made use of an independent validation set collected as part of the \gls{CGLS-LC100} project \citep{nandin-erdene_tsendbazar_copernicus_2019} that includes a total of over 20\,000 points collected across the globe, using the same legend, grid and spatial resolution as in our study.
The validation data was collected at \gls{WUR}, whereas the training data was collected at \gls{IIASA}, ensuring independence of the datasets.
We used the validation data to assess model accuracy both thematically and spatially.

All in all, with this study we aim to investigate and compare methods for performing global land cover fraction mapping, and with a comprehensive set of methods as well as a wide variety of input covariates, so as to improve land cover fraction map production.
We also aim to suggest ways of dealing with the balance issues inherent to land cover fraction reference data that hampers land cover fraction mapping, and to assess the accuracy of the result.
Therefore the objectives of our study are:

\begin{enumerate}
 \item Compare the performance of a variety of machine learning regression models for global-scale land cover fraction mapping
 \item Investigate approaches for tuning the models by reducing bias in the predictions with regards to zero inflation and predictions tending towards the mean
 \item Compare the importance of covariates and covariate groups for the prediction of land cover fractions
 \item Assess the accuracy of the models from a thematic and spatial point of view
\end{enumerate}

@@

The reference data (for model training and validation) used in this study was collected as part of the \ac{CGLS-LC100} project \citep{buchhorn_copernicus_2020, nandin-erdene_tsendbazar_copernicus_2019}.
The training set was generated by experts in \ac{IIASA}.
This global point dataset provides information on nine class fractions of sampled pixels of the PROBA-V 100 m product \citep{dierckx2014probav}.
The fraction estimate was acquired by high-resolution image interpretation by experts, for the year 2015.
Each of the sampled PROBA-V pixels was subdivided into a 10-by-10 m subpixel grid and the class fraction estimates were derived from the subpixel ratios.

For model validation, an independent global dataset of \ac{CGLS-LC100}, generated by \gls{WUR} \citep{tsendbazar_developing_2018}, was used.
The method of point collection is equivalent to the method used for training data collection, but performed independently by a different group of regional experts.

@@

%Since the data describes fractions, it is zero-inflated, since it is relatively rare for more than just a few land cover classes to occupy space within an area of a single pixel.
%This means that for a given pixel, only a few of the classes would normally have a fraction that is above 0\%.
%This leads to a dataset that is zero-inflated (also known as sparse), with over half of all values for a pixel being zero.
%Such an imbalance is a problem for statistical and machine learning techniques that are tuned by minimising the error of the output, since they are heavily biased towards a good prediction of zeroes.
%In other words, when validating against a zero-inflated dataset, a prediction of zero is a safe bet.
%But for user communities, correct prediction of non-zero values is more important.

@@

To predict land cover fractions in unsampled areas, the models learn from covariates present at the locations of the training data, and then the covariates are used for prediction in unsampled areas.
See figure \ref{fig-processing} for an overview of the whole processing chain.

In this study, six groups of covariates were used: vegetation indices, temporal metrics, terrain metrics, soil metrics, climate metrics and location data (see table \ref{tab-inputdata}).
All of these covariates had to be preprocessed before they could be input into the models.

\begin{figure}
 \centering
 \includegraphics[width=0.75\textwidth]{article-figures/flowcharts/2020-07-10-flowchart}
 \caption{Processing workflow, from the raw input data to model accuracy assessment and wall-to-wall map output.}
 \label{fig-processing}
\end{figure}

@@

%\subsubsection{Vegetation indices and temporal metrics}
\label{sec-temporal-filter}

The entire archive (2014-03-11 to 2019-07-16) of the PROBA-V 100 m Level 3 Top-of-Canopy 5-day composite product \citep{dierckx2014probav,probavguide2} was used for this study.
While the reference data corresponded to the land cover at the year 2015 specifically, the whole time series of PROBA-V was chosen to obtain more robust statistics for the temporal metrics.
Land cover change is a relatively rare phenomenon and thus, according to tests performed in the making of \gls{CGLS-LC100} collection 3, is not likely to affect more than 2\% of the training points.
Nevertheless, this may result in poorer model performance compared to if reference data for each year would be available.

The time series of each of the PROBA-V spectral bands were cloud masked, first by applying the status mask provided with the product itself, and then by running a temporal cloud filter to remove the remaining outliers that were further than 2 standard deviations away from the fitted \ac{LOESS} curve over the blue reflectance band.
The resulting cloud-free time series was then used to generate the following \glspl{VI}: \gls{NDVI}, \gls{NDMI}, \gls{EVI}, \gls{OSAVI} and \gls{NIRv}. Next, the \glspl{VI} were used to calculate the descriptive statistics (median and interquartile range) over the whole time series, as well as for each phenological season separately.
Next, harmonic analysis \citep{jakubauskas2001harmonic} was run on NDVI in order to decompose the time series into sine and cosine components for two frequency orders (annual and semiannual), as well as the trend and intercept of the model.
From this, phase and amplitude for the two orders were calculated.
Lastly, the minimum and maximum of the NDVI time series was also retained as covariates.

%\subsubsection{Terrain covariates}

For elevation data, the ASTER GDEM v003 \citep{ASTGTM003} product (30 m) was resampled to PROBA-V 100 m grid and used as the elevation covariate.
In addition, \gls{GDAL} \citep{gdal} was used to calculate terrain parameters out of elevation: slope, aspect and \ac{TPI}.
%All terrain covariates were then resampled to the PROBA-V grid.

%\subsubsection{Climate covariates}

@@

%\subsubsection{Soil covariates}

@@

%\subsubsection{Location covariates}

@@

In total, 313 covariates were generated in pre-processing.
However, many of them are collinear with one another, which prolongs training time for machine learning models and leads to unreliable coefficient estimation and increased error variance for linear models.
Thus, variable selection was used to remove collinear covariates before predicting the land cover fractions.
Covariates that had a Pearson's correlation coefficient $r$ \citep{pearson_notes_1895} above 0.9 were excluded in an iterative process.
After that, covariates with a Spearman's rank correlation $\rho$ \citep{spearman1904rank} above 0.9 were likewise excluded.

Most of the collinear covariates were soil covariates at different depths.
Therefore the 10 cm depth covariates were left in, and the other depths excluded, as long as $r$ was above 0.9.
Similarly, climate data was collinear between subsequent months. Thus January and July data was preferred, as these months represent different extremes of the year and are less correlated with other months' data.
In addition, all of the spectral bands of PROBA-V were also highly correlated with each other as well as to \glspl{VI}, so only \glspl{VI} were kept.

After the covariate selection process, 67 covariates remained.
These covariates include data from each of the covariate categories.
See table \ref{tab-inputdata} for an overview.

%\begin{figure}
 %\inputdigraph[width=\textwidth]{article-figures/algorithms}{dot}
% \includegraphics[width=\textwidth]{article-figures/flowcharts/algorithms}
% \caption{Preprocessing chain.}
% \label{fig-preprocessing}
%\end{figure}

@@

For deriving land cover fractions, we compared a wide array of machine learning regression methods.
We broadly divided the methods into three types: linear models, machine learning models based on \glspl{CART}, and machine learning models not based on \glspl{CART}.
See table \ref{tab-methods} for the full list of methods that were compared in this study.
%They can be broadly divided into three categories.
%Tested linear models were \ac{FNC}, \ac{GLM}, \ac{PLS} regression, Lasso regression and \ac{MLR}.
%Non-tree machine learning models tested were \ac{MLP} \glspl{NN} and \glspl{SVM}.
%Lastly, machine learning models based on \glspl{CART} tested were Cubist and \ac{RF} regressions.
In addition, as a baseline we also compared the results with an intercept model (all fractions always predicted to be equal, namely $\frac{100\%}{7}\approx14.29\%$).

All of the models were made using the free and open-source statistical software R \citep{r_2019}.

@@

\subsubsection{Linear models}
%\subsubsection{\Glsentryfull{GLM}}

\Glsentryfull{GLM} \citep{neter_applied_1996}, also known as multivariate linear regression, is an extension to a regular linear regression that allows for multiple outcomes.
In this study, we used \gls{GLM} as a baseline model.
We also tested whether dropping any additional covariates from the model would result in better predictions, using a sphericity test with Greenhouse-Geisser correction to compare models with different covariates.
However, only three covariates could be dropped and it did not improve the predictions, therefore we decided to keep all the covariates in the model.
\Gls{GLM} is implemented in base R \citep{r_2019}.

%\subsubsection{Lasso/ridge/elastic net regression}

>>>>

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lineno}
\usepackage{microtype}
\usepackage{amsmath}
%\usepackage{indentfirst}
\usepackage{booktabs} % For \toprule, etc.
\usepackage{textcomp} % for degree sign
%\usepackage[sort&compress,sectionbib]{natbib} % already in elsarticle
\usepackage{graphicx}
%\usepackage{adjustbox} % For resizing tables
\usepackage{array} % For mid-alignment (m{xcm}) in tables
\usepackage{rotating} % Sideways figures
%\bibliographystyle{apalike} % elsarticle defined at the top

@@

% Put table captions on top
\usepackage{floatrow}
\floatsetup[table]{capposition=top}

% My own definition for the sections that start with a bold name but don't appear in TOC
\newcommand{\minisection}[1]{\paragraph{#1}}%{\medskip \textbf{#1:}}
% Package names
\newcommand{\cran}[1]{CRAN: #1}

@@

% Acronyms - cite with \ac{}
\newacronym{GNSS}{GNSS}{Global Navigation Satellite System}
\newacronym{OA}{OA}{overall accuracy}
\newacronym{UA}{UA}{user accuracy}
\newacronym{PA}{PA}{producer accuracy}
\newacronym{RMSE}{RMSE}{root mean squared error}
\newacronym{MAE}{MAE}{mean absolute error}
\newacronym{ME}{ME}{mean error}
\newacronym{NSE}{NSE}{Nash–Sutcliffe model efficiency coefficient}
\newacronym{SDG}{SDG}{Sustainable Development Goal}
\newacronym{LC-CCI}{LC-CCI}{ESA Climate Change Initiative Land Cover}
\newacronym{CGLS-LC100}{CGLS-LC100}{Copernicus Global Land Services 100 m Land Cover}
%\newacronym{CGLS}{CGLS}{Copernicus Global Land Services}
\newacronym{FROM-GLC}{FROM-GLC10}{Finer Resolution Observation and Monitoring Global Land Cover}
\newacronym{IIASA}{IIASA}{International Institute for Applied Systems Analysis}
\newacronym{WUR}{WUR}{Wageningen University \& Research}
\newacronym{SMA}{SMA}{spectral mixture analysis}
\newacronym{SVM}{SVM}{support vector machine}
\newacronym{MLP}{MLP}{multi-layer perceptron}
\newacronym{NN}{NN}{neural network}
\newacronym{LOESS}{LOESS}{locally estimated scatterplot smoothing}
\newacronym[firstplural={vegetation indices (VIs)}]{VI}{VI}{vegetation index}
\newacronym{GLSDEM}{GLSDEM}{Global Land Survey Digital Elevation Model}
\newacronym{TPI}{TPI}{Terrain Position Index}
\newacronym{OSAVI}{OSAVI}{Optimised Soil-Adjusted Vegetation Index}
\newacronym{NDVI}{NDVI}{Normalised Difference Vegetation Index}
\newacronym{NDMI}{NDMI}{Normalised Difference Moisture Index}
\newacronym{NIRv}{NIR\textsubscript{v}}{\gls{NIR} of vegetation}
\newacronym{EVI}{EVI}{Enhanced Vegetation Index}
\newacronym{NIR}{NIR}{near infra-red}
\newacronym{SIF}{SIF}{sun-induced chlorophyll fluorescence}
\newacronym{GPP}{GPP}{gross primary productivity}
\newacronym{MSI}{MSI}{multi-spectral instrument}
\newacronym{SAR}{SAR}{synthetic aperture radar}
\newacronym{SCM}{SCM}{subpixel confusion-uncertainty matrix}
\newacronym{FNC}{FNC}{fuzzy nearest centroid}
\newacronym{PLS}{PLS}{partial least squares}
\newacronym{MLR}{MLR}{multinomial logistic regression}
\newacronym{RF}{RF}{random forest}
\newacronym{TOC}{TOC}{Top-of-Canopy}
\newacronym{CART}{CART}{classification and regression tree}
\newacronym{GLM}{GLM}{general linear regression model}
\newacronym{LCCS}{LCCS}{UN Land Cover Classification Scheme}
\newacronym{GDAL}{GDAL}{Geospatial Data Abstraction Library}
%\makenoidxglossaries
% Disable abbreviating RF for now [text={random forest}]
%\glsunset{RF}

@@

\begin{abstract} % elsarticle: environment instead of a command \abstract{
Currently most global land cover maps are produced with discrete classes, which express the dominant land cover class in each pixel, or a combination of several classes at a predetermined ratio. In contrast, land cover fraction mapping allows for expressing the proportion of each pure class in each pixel, which increases precision and reduces legend complexity. To map land cover fractions, regression rather than classification algorithms are needed, and multiple approaches are available for this task.

@@

One challenge for models for land cover fraction mapping is data sparsity: the dataset is zero-inflated, but regression favours the mean, thus 0\% and 100\% are difficult to predict well. We proposed a new solution by combining three models. A binary model determines whether a pixel is pure; if so, it is processed using a classification model; otherwise with a regression model.

@@

@@

Land cover, as one of the key variables for monitoring a number of \glspl{SDG}, has received increasing amounts of attention, also as new satellites are launched and the capacity for land monitoring increases.
In this context, new global land cover maps have emerged to better map the current land cover of the world, as well as to track land cover change.
Some of the recent achievements have been the \ac{LC-CCI} map \citep{esa_land_2017} that aims at producing a long-term set of consistent global annual medium-resolution land cover maps aimed at the climate community, \glsentryfull{CGLS-LC100} map \citep{buchhorn_moderate_2019, buchhorn_copernicus_2020} that aims at a finer spatial resolution and high quality with yearly updates since 2015, and the \ac{FROM-GLC} map \citep{fromglc2019} that showcases the potential of land cover mapping at 10 m resolution.

Except for the \ac{CGLS-LC100} product with its cover fraction layers, all other global land cover maps, including previous maps such as by \citet{bartholome2005glc2000, friedl2010modis, arino2007globcover, see2015hybrid, chen2015globeland30}, are provided with discrete classes (also known as ``hard'' or ``crisp'' classification), where each pixel of the map can only represent a single class.
However, this results in less precise estimation of land cover due to mixed pixels, where in reality there are multiple land cover classes in the area covered by a single pixel.
This issue becomes worse at coarse resolutions and heterogeneous landscapes.
It may result in biases, for instance, a sparse forest may be classified as grassland, ignoring the relatively few trees in the area, and thus underestimate tree cover in the pixel.
These systematic errors then add up when scaling the result to the entire region.

A potential solution to this issue is to characterise land cover using cover fractions.
In this approach, instead of a discrete class, the proportion of every class in the legend is reported for every pixel of the map.
That way, the land cover models work not on pixel labels, such as ``forest'', but on land cover characteristics, such as tree cover, defined as the area of the pixel covered by tree canopies, as well as herbaceous cover, defined as the area not covered by woody vegetation.
This is also called ``fuzzy'' or ``soft'' classification, and sometimes ``subpixel'' classification or ``linear mixture modelling'' \citep{Okeke2006fuzzyexponent}.
Related terms are ``super-resolution mapping'' and ``subpixel mapping'', but these concepts include the allocation of land cover fractions inside a pixel spatially, rather than just determining the fraction ratio per pixel.

Land cover fraction mapping has been attempted in the past.
Most of the previous research has focused on deriving land cover fractions of 3-6 classes at a local scale  \citep{adams_classification_1995, foody1996fuzzyevaluation, walton2008subpixelrf, hansen_continuous_2011, sharma_assessing_2011, uma_shankar_wavelet-fuzzy_2011, dwivedi_optimisation_2012, lizarazo_quantitative_2012, gessner_estimating_2013, okujeni_generalizing_2018}, less often at a regional scale and with more detailed classes \citep{colditz_land_2011}.
The methods for assessing the accuracy of the results vary greatly between the different studies.
In addition, global land cover fraction products have emerged, but focused on a particular class, such as tree cover \citep{hansen_global_2003, Hansen2013forestchange}, water \citep{schroeder_development_2015} and urban area \citep{gao_mapping_2020}.
To date only \ac{CGLS-LC100} \citep{marcel_buchhorn_copernicus_2019} provides global maps with fractions of every major land cover class \citep{nandin-erdene_tsendbazar_copernicus_2019}.

Land cover fraction mapping can be performed using a variety of different approaches and algorithms.
In its core, it is a regression rather than a classification problem, as the output is a fraction of a label rather than a label itself.
Popular approaches include fuzzy nearest centroid regression \citep{zhang2001fullyfuzzy}, \ac{SMA} (e.g. \citealp{shimabukuro1991least, adams_classification_1995, hobbs2003linear, yang_landsat_2012}), \ac{RF} regression \citep{walton2008subpixelrf}, \gls{SVM} regression \citep{walton2008subpixelrf}, Cubist regression \citep{walton2008subpixelrf}, \gls{MLP} \glspl{NN} \citep{zhang2001fullyfuzzy}, genetic algorithms \citep{stavrakoudis_boosted_2011} and wavelet transformation \citep{uma_shankar_wavelet-fuzzy_2011}.
So far other studies have only compared a few of these methods at once, and never with a thematically complete set of land cover classes nor at global scale.

In addition to classification methods, one aspect that is important for classification accuracy is the data (covariates) that the algorithm is trained on and predict land cover from \citep{yu2014metadiscoveries}. In addition to optical covariates (such as blue, red, \gls{NIR} and shortwave infrared spectral bands), temporal covariates (growing season start, duration, growing intensity) are commonly used to improve the separability of vegetation, such as crop types \citep{jakubauskas2001harmonic}. Elevation covariates (elevation, slope, aspect) are known to improve the separability between classes, such as trees \citep{burrough2001fuzzy}. \Glspl{VI} condense data and reduce the noise inherent in single bands, and thus are often used to improve land cover class separability, e.g. for wetlands \citep{sader1995wetlands}, as well as for defining classification rules in rule-based fuzzy classification \citep{baraldi2006rulebased}. Therefore in this paper such covariates have also been used in order to increase the accuracy of land cover fraction regression model predictions, and their importance for the predictions was analysed.

A common issue with the use of land cover fraction data as input into regression models is data imbalance.
The more classes are mapped, the more often there will be 0\% of a given class in any particular pixel, leading to zero inflation.
In addition, in homogeneous areas, 100\% cover of a single class is common.
This leaves little data in the middle for the regression models to learn from.
Conversely, regression models tend to favour predictions closer to the mean and rarely predict extreme values.
We investigated this issue and suggested ways to deal with it.

Lastly, accuracy assessment by itself is often times challenging, especially at a global scale.
It requires a comprehensive dataset across the globe that would be comparable with the modelled land cover, and yet independent of the training data used to create it.
For land cover fraction mapping it is even more challenging, as fraction information needs to be available, or alternatively fine spatial resolution data from which the fractions can be calculated.
Because of these challenges, the information about the accuracy of the existing global land cover fraction products is often limited, which in turn is problematic for the users to decide on what product suits their needs the best.
Users would also benefit from knowing the spatial variation of accuracy, as models may be more accurate at certain locations of the world and less so at others.
In this study, we made use of an independent validation set collected as part of the \gls{CGLS-LC100} project \citep{nandin-erdene_tsendbazar_copernicus_2019} that includes a total of over 20\,000 points collected across the globe, using the same legend, grid and spatial resolution as in our study.
The validation data was collected at \gls{WUR}, whereas the training data was collected at \gls{IIASA}, ensuring independence of the datasets.
We used the validation data to assess model accuracy both thematically and spatially.

All in all, in this study we investigated and compared methods for performing global land cover fraction mapping, and with a comprehensive set of methods as well as a wide variety of input covariates, so as to improve land cover fraction map production.
We also investigated ways of dealing with the balance issues inherent to land cover fraction reference data that hampers land cover fraction mapping, and to assess the accuracy of the result.
Therefore the objectives of our study were to:

\begin{enumerate}
 \item Compare the performance of a variety of machine learning regression models for global-scale land cover fraction mapping.
 \item Investigate approaches for tuning the models by reducing bias in the predictions with regards to zero inflation and predictions tending towards the mean.
 \item Compare the importance of covariates and covariate groups for the prediction of land cover fractions.
 \item Assess the accuracy of the models from a thematic and spatial point of view.
\end{enumerate}

@@

The reference data (for model training and validation) used in this study was collected as part of the \ac{CGLS-LC100} project \citep{buchhorn_copernicus_2020, nandin-erdene_tsendbazar_copernicus_2019}.
The training set was generated by experts in \ac{IIASA}.
This global point dataset provides information on fractions of each class within sampled pixels of the PROBA-V 100 m product \citep{dierckx2014probav}.
The fraction estimates were acquired by high-resolution image interpretation by experts, for the year 2015.
Each of the sampled PROBA-V pixels was subdivided into a 10-by-10 m subpixel grid and the class fraction estimates were derived from the subpixel ratios.

For model accuracy assessment, an independent global validation dataset of \ac{CGLS-LC100}, generated by \gls{WUR} \citep{tsendbazar_developing_2018}, was used.
The method of point collection is equivalent to the method used for training data collection, but performed independently by a different group of regional experts.

@@

@@

To predict land cover fractions in unsampled areas, the models learn from covariates present at the locations of the training data, and then the covariates are used for prediction in unsampled areas.
See figure \ref{fig-processing} for an overview of the whole processing chain.
The processing was carried out in R \citep{r_2019} and the resulting code has been made openly available in \citet{dainius_masiliunas_2020_3973123}.

In this study, six groups of covariates were used: vegetation indices, temporal metrics, terrain metrics, soil metrics, climate metrics and location data (see supplementary material, table S1).
These covariates had to be preprocessed before they could be input into the models.

\begin{figure}
 \centering
 \includegraphics[width=0.6\textwidth]{article-figures/flowcharts/2020-07-10-flowchart}
 \caption{Processing workflow, from the raw input data to model accuracy assessment and wall-to-wall map output.}
 \label{fig-processing}
\end{figure}

@@

\label{sec-temporal-filter}

The entire archive (2014-03-11 to 2019-07-16) of the PROBA-V 100 m Level 3 Top-of-Canopy 5-day composite product \citep{dierckx2014probav,probavguide2} was used for this study.
While the reference data corresponded to the land cover at the year 2015 specifically, the whole time series of PROBA-V was chosen to obtain more robust statistics for the temporal metrics.
Land cover change is a relatively rare phenomenon, according to tests performed in the making of \gls{CGLS-LC100} collection 3, and so we expected that the difference caused by land cover change to the input data would be smaller than the margin of error of the model output.
Nevertheless, this may result in poorer model performance compared to if up-to-date reference data for each year would be available.

The time series of each of the PROBA-V spectral bands were cloud masked, first by applying the status mask provided with the product itself, and then by running a temporal cloud filter to remove the remaining outliers that were further than 2 standard deviations away from the fitted \ac{LOESS} curve over the blue reflectance band.
The resulting cloud-free time series was then used to generate the following \glspl{VI}: \gls{NDVI}, \gls{NDMI}, \gls{EVI}, \gls{OSAVI} and \gls{NIRv}. Next, the \glspl{VI} were used to calculate the descriptive statistics (median and interquartile range) over the whole time series, as well as for each phenological season separately.
Next, harmonic analysis \citep{jakubauskas2001harmonic} was run on NDVI in order to decompose the time series into sine and cosine components for two frequency orders (annual and semiannual), as well as the trend and intercept of the model.
From this harmonic analysis we obtained minimum and maximum values of the \gls{NDVI} time series, and calculated the phase and amplitude for the two harmonic orders from the respective sine and cosine components.

For elevation data, the ASTER GDEM v003 \citep{ASTGTM003} product (30 m) was resampled to PROBA-V 100 m grid and used as the elevation covariate.
In addition, \gls{GDAL} \citep{gdal} was used to calculate terrain parameters out of elevation: slope, aspect and \ac{TPI}.

@@

@@

@@

In total, 313 covariates were generated in pre-processing.
However, many of them were collinear with one another.
Multicollinearity prolongs training time for machine learning models and leads to unreliable coefficient estimation and increased error variance for linear models.
Thus, variable selection was used to remove collinear covariates before predicting the land cover fractions.
Covariates that had a Pearson's correlation coefficient $r$ \citep{pearson_notes_1895} above 0.9 were excluded in an iterative process.
After that, covariates with a Spearman's rank correlation $\rho$ \citep{spearman1904rank} above 0.9 were likewise excluded.

The covariates to exclude were selected manually, to avoid interpretation difficulties that arise from automatic selection procedures.
The majority of the collinear covariates were soil metrics predicted at different depths.
Therefore the 10 cm depth covariates were left in, and the other depths excluded, as long as $r$ was above 0.9.
Similarly, climate data was collinear between subsequent months. Thus January and July data was preferred, as these months represent different extremes of the year and are less correlated with other months' data.
Our initial correlation analysis also showed that the spectral bands of PROBA-V were highly correlated with each other as well as to \glspl{VI}, so only \glspl{VI} were used as covariates.

After the covariate selection process, 67 covariates remained.
These covariates include data from each of the covariate categories.
See supplementary material, table S1 for an overview.

@@

For deriving land cover fractions, we compared a wide array of machine learning regression methods.
We broadly divided the methods into three types: linear models, machine learning models based on \glspl{CART}, and machine learning models not based on \glspl{CART}.
We tuned the methods to select optimal parameters for each.
See table \ref{tab-methods} for the full list of methods that were compared in this study.
In addition, as a baseline we also compared the results with an intercept model (all fractions always predicted to be equal, namely $\frac{100\%}{7}\approx14.29\%$).
All of the models were made using the free and open-source statistical software R \citep{r_2019}.

@@

\subsubsection{Linear models}

The \gls{GLM} \citep{neter_applied_1996}, also known as multivariate linear regression, is an extension to the standard linear regression that allows for multiple outcomes.
In this study, we used a \gls{GLM} as a baseline model.
We also tested whether dropping any additional covariates from the model would result in better predictions, using a sphericity test with Greenhouse-Geisser correction \citep{greenhouse_methods_1959} to compare models with different covariates.
However, only three covariates could be dropped and it did not improve the predictions, therefore we decided to keep all the covariates in the model.
\Gls{GLM} is implemented in base R \citep{r_2019}.
